---
title: "Airway Microbiome Analysis"
author: "Dan Kerchner"
date: "2023-05-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
```
# Load and merge data
```{r}
# MICROBIOME DATA
md <- read.csv('data/metadata_nose_v2.csv', blank.lines.skip = TRUE)
md <- md %>%
  select(-c(SRAcode, patient, season))

# CLINICAL DATA
rd <- read.csv('data/Registo_working.csv')
rd <- rd %>%
  mutate(Gender = case_when(Gender == "Feminino" ~ "F",
                            Gender == "Masculino" ~ "M")) %>%
  mutate(season = case_when(season == "spring" ~ "spring",
                             TRUE ~ "not spring"))

# JOIN
df <- left_join(md, rd, by=c('ID', 'asthma_rhinitis')) 

df <- df %>%
  select(-ID) %>%
  mutate(asthma_rhinitis = factor(asthma_rhinitis, levels = c("AR", "ARAS")),
         TCUT_Positive_IgE_for_Aeroallergens = factor(TCUT_Positive_IgE_for_Aeroallergens),
         Bronchodilation_Positive = factor(Bronchodilation_Positive))

# Omit incomplete observations
df_all <- na.omit(df)

# Renumber
rownames(df_all) <- 1:nrow(df_all)

# Microbiome data, but only for the patients who also had complete clinical data
microbiome_df <- df_all %>%
  select(asthma_rhinitis, Firmicutes:Prevotella)
```

# Logistic regression

```{r}
set.seed(1234)

model1 <- glm(asthma_rhinitis ~ .,
              family = binomial(link = "logit"),
              data = df_all)

# 10-fold CV for logistic regression
model1_logit_cv <- train(x = df_all[2:length(names(df_all))],
                         y = df_all$asthma_rhinitis,
                         method = "glm", family = binomial(link = "logit"),
                         trControl = trainControl(method = "cv", number = 10,
                                                  summaryFunction = twoClassSummary,
                                                  classProbs = TRUE, savePredictions = TRUE))
model1_logit_cv$results
```

```{r}
fm1 <- model1_logit_cv$finalModel
summary(fm1)
```

```{r message=FALSE, warning=FALSE}
fm1_coefs <- coef(fm1)[-1]
fm1_names <- rownames(fm1$R)[-1]
fm1_confint <- confint(fm1)[-1, ]
fm1_results <- data.frame(fm1_names, fm1_coefs,
                          fm1_ll = fm1_confint[,1],
                          fm1_ul = fm1_confint[,2]) %>%
  arrange(desc(fm1_coefs))

ggplot(fm1_results) +
  geom_point(aes(y = fm1_names, x = fm1_coefs)) + 
  geom_errorbar(aes(y = fm1_names, x = fm1_coefs,
                    xmin=fm1_ll, xmax=fm1_ul))
```

```{r}
prob_thresh <- seq(0.05, 0.95, by = 0.05)
model1_logit_ths <- thresholder(model1_logit_cv, threshold = prob_thresh,
                                statistics = c("Sensitivity", "Specificity", "Accuracy", "Kappa"))
model1_logit_ths
```

We select the classification threshold by maximizing Cohen's $\kappa$.

```{r}
model1_threshold <- model1_logit_ths[which.max(model1_logit_ths$Kappa), ]
model1_threshold
```

Where Kappa is maximized, at a threshold of 0.70, the test accuracy is 0.6802.

# Stepwise Selection

```{r}
model2_stepwise <- step(model1, direction = "both",
                        k = log(nrow(df_all)), # This k gives BIC
                        trace = FALSE) 
model2_stepwise$formula
```
```{r}
# Which variables are still in?
model2_xnames <- names(model2_stepwise$coefficients)[-1]
model2_xnames
```

## Manually estimate test accuracy using 10-fold cross-validation 
```{r}
library(psych)
set.seed(12345)
mean_accuracy <- 0.0
n_cv <- 10
for(i in 1:n_cv){
  df_sample_train_idx <- sample(1:nrow(df_all), size = 0.75*nrow(df_all))
  df_sample_train <- df_all[df_sample_train_idx,]
  df_sample_test <- df_all[-df_sample_train_idx,]
  model2_i <- glm(asthma_rhinitis ~ .,
                  family = binomial(link = "logit"),
                  data = df_sample_train)
  model2_stepwise_cv_i <- step(model1,
                               direction = "both",
                               k = log(nrow(df_all)),
                               trace = 0)
  prob_thresh <- seq(0.1, 0.9, by = 0.1)

  # We select the classification threshold by maximizing Cohen's $\kappa$.
  max_k <- 0
  best_pt <- NA
  accuracy <- 0
  for(pt in prob_thresh) {
    model2_test_pred_class <- factor(ifelse(
      predict(model2_stepwise_cv_i, newdata = df_sample_test[,-1], type = "response") >= pt,
      "ARAS",
      "AR"),
      levels = c("AR", "ARAS"))
    
    ck_i <- cohen.kappa(cbind(df_sample_test$asthma_rhinitis, model2_test_pred_class))
    if(ck_i$kappa >= max_k){
      max_k <- ck_i$kappa
      best_pt <- pt  # Not sure why this is always coming out to 0.5
      accuracy <- sum(df_sample_test$asthma_rhinitis == model2_test_pred_class)/length(df_sample_test)
      #print(accuracy)
    }
  }
  mean_accuracy <- mean_accuracy + accuracy/n_cv
}
print(mean_accuracy)
```

Mean accuracy as estimated by 10-fold cross-validation is `r mean_accuracy`.




# LASSO Regression

```{r}
########
# Lasso
########
library(glmnet)
# LASSO regression with glmnet (from glmnet package)
# the glmnet() function fits the "complete" solution paths
df_y <- ifelse(df_all$asthma_rhinitis == "AR", "0", "1")
df_x <- model.matrix(asthma_rhinitis ~ ., data = df_all)[, -1]
df_lasso <- glmnet(x = df_x,
                   y = df_y,
                   alpha = 1,
                   family = 'binomial')
lassologit_tune_grid <- data.frame(alpha = 1, lambda = df_lasso$lambda)
# plot solutions paths (from plotmo package)
plot(df_lasso, xvar = "lambda")

# use 10-fold CV to select the tuning parameter lambda
set.seed(1234)
df_lasso_cv <- cv.glmnet(x = df_x,
                         y = df_y,
                         method = "glmnet",
                         family = "binomial",
                         tuneGrid = lassologit_tune_grid,  # Using this resulted in all variables dropping out
                         alpha = 1, nfolds = 10,
                         type.measure = 'class')

# Find index of minimum lambda
min_lambda_idx <- which(df_lasso_cv$lambda==df_lasso_cv$lambda.min)

# Compute accuracy at that index, as 1 minus cross validation error
df_lasso_cv_accuracy = 1 - df_lasso_cv$cvm[min_lambda_idx]
```

Cross-validation accuracy for LASSO is `r df_lasso_cv_accuracy`

```{r}

plot(df_lasso_cv$glmnet.fit, xvar = "lambda")
```

```{r}

# look at solution corresponding to selected lambda based on minimum 10-fold CV MSE
# note that this model is fit on the entire training data set
predict(df_lasso_cv, s = df_lasso_cv$lambda.min, type = "coef")

```

```{r}
##### 
# Group lasso using gglasso
#####

# library(gglasso)
# set.seed(1234)
# group_num <- c(1:35, 36, 36, 36) # group 36 is for "season"
# model_gl <- gglasso(x = df_x,
#                     y = factor(df_all$asthma_rhinitis, levels = c("AR", "ARAS")),
#                     group = group_num)

# Results in the following, which seems to contradict the documentation that states that
# for logistic regression, y should be a factor.
# 
# Error in gglasso(x = df_x, y = factor(df_all$asthma_rhinitis, levels = c("AR",  : 
#   The response y must be numeric. Factors must be converted to numeric

```

# Random Forest

```{r}
library(randomForest)

####
# randomForest with full model
####
set.seed(1234)
rf_train <- train(x = df_all %>% select(-asthma_rhinitis),
                   y = df_all$asthma_rhinitis,
                   method = "rf", ntree = 500,
                  tuneGrid = data.frame(mtry = 1:10),
                  trControl = trainControl(method = "cv", number = 10))
rf_train

# Confirm that 500 trees is more than enough
plot(x = 1:500, y = rf_train$finalModel$err.rate[,1], type = "l",
     xlab = "Number of Trees", ylab = "OOB Error Rate")
```

```{r}
# obtain OOB error rate / accuracy for selected model
rf_train$results$Accuracy[rf_train$bestTune$mtry]
```

```{r}
varImpPlot(rf_train$finalModel, main = "Variable Importance Plot", n.var = 10)
```

# Random Forest with reduced model
```{r}
####
# randomForest with reduced model, result of stepwise selection
####
set.seed(1234)
rf_train <- train(x = df_all %>% select(Firmicutes, Proteobacteria, Actinobacteriota, Fusobacteriota,
                                        Bacteroidota, Moraxella, Fev1.post, MMEF75_25.post,
                                        Nitric_Oxide, season),
                   y = df_all$asthma_rhinitis,
                   method = "rf", ntree = 500,
                  tuneGrid = data.frame(mtry = 1:10),
                  trControl = trainControl(method = "cv", number = 10))
rf_train
# Confirm that 500 trees is more than enough
plot(x = 1:500, y = rf_train$finalModel$err.rate[,1], type = "l",
     xlab = "Number of Trees", ylab = "OOB Error Rate")
```

```{r}
# obtain OOB error rate / accuracy for selected model
rf_train$results$Accuracy[rf_train$bestTune$mtry]
```
```{r}
varImpPlot(rf_train$finalModel, main = "Variable Importance Plot", n.var = 10)
```


